---
title: "Tetuoan City Power Consumption"
author: "Alyssa Mercado"
format: html
editor: visual
execute:
  working-directory: ./
---

Uses various regression models to determine the cause of usage and predict future consumption

## Cleaning and splitting the data

After reading the CSV a seed is used for reproducibility. The date and time predictors are extracted to be DateTime variables and turned into unordered factors Data is split 75/25 into training and testing data sets

```{r}
#| echo: false
#| message: false
#| warning: false

# libraries 
library(lubridate)
library(dplyr)
library(glmnet)
library(randomForest)
library(caret)
library(pdp)

library(xgboost)

tetouan <- read.csv("dataset/Tetuan City power consumption.csv", header = TRUE)
set.seed(20260429)

# extract dates and times
tetouan$DateTime <- as.POSIXct(tetouan$DateTime, format = "%m/%d/%Y %H:%M") # //  
tetouan$Hour   <- factor(hour(tetouan$DateTime), ordered = FALSE)
tetouan$Weekday <- factor(wday(tetouan$DateTime, label = TRUE), ordered = FALSE)
tetouan$Month  <- factor(month(tetouan$DateTime, label = TRUE), ordered = FALSE)

tetouan <- na.omit(tetouan)
colnames(tetouan) <- c(
  "DateTime", "Temperature", "Humidity", "WindSpeed",
  "GeneralDiffuse", "DiffuseFlows",
  "Zone1", "Zone2", "Zone3",
  "Hour", "Weekday", "Month"
)


# train & test data
n <- dim(tetouan)[1]
trainind <- sample(1:n, round(n*0.75))
testind <- c(1:n)[-trainind]

train <- tetouan[trainind, ]
test <- tetouan[testind, ]

```

## Exploratory Data Analysis

```{r}
#| echo: false
#| fig.show: hold
#| fig.cap: "Distribution of power consumption in each zone"

# general amount of power consumption per zone

hist(tetouan$Zone1, main = "Zone 1 Power Consumption", xlab = "Power Consumption", col = "lightblue")
hist(tetouan$Zone2, main = "Zone 2 Power Consumption", xlab = "Power Consumption",  col = "lightgreen")
hist(tetouan$Zone3, main = "Zone 3 Power Consumption", xlab = "Power Consumption", col = "salmon")
```

```{r}
#| echo: false
#| fig.show: hold
#| fig.cap: "Monthly power consumption per zone"

boxplot(Zone1 ~ Month, data = tetouan, main = "Zone 1 by Month", xlab = "Month", ylab = "Power", col = "lightblue")
boxplot(Zone2 ~ Month, data = tetouan, main = "Zone 2 by Month", xlab = "Month", ylab = "Power", col = "lightgreen")
boxplot(Zone3 ~ Month, data = tetouan, main = "Zone 3 by Month", xlab = "Month", ylab = "Power", col = "salmon")

```

```{r}
#| echo: false
#| fig.show: hold
#| fig.cap: "Weekly power consumption per zone"

# power consumption by weekday
boxplot(Zone1 ~ Weekday, data = tetouan, 
        main = "Zone 1 by Weekday", xlab = "Weekday", ylab = "Power", col = "lightblue")
boxplot(Zone2 ~ Weekday, data = tetouan, 
        main = "Zone 2 by Weekday", xlab = "Weekday", ylab = "Power", col = "lightgreen")
boxplot(Zone3 ~ Weekday, data = tetouan, 
        main = "Zone 3 by Weekday", xlab = "Weekday", ylab = "Power", col = "salmon")

```

```{r}
#| echo: false
#| fig.show: hold
#| fig.cap: "mean consumption by hour for each zone"


#### mean consumption by hour for each zone
hourly_mean <- tetouan %>%
  group_by(Hour) %>%
  summarise(
    mean_zone1 = mean(Zone1),
    mean_zone2 = mean(Zone2),
    mean_zone3 = mean(Zone3)
  )

# convert hour from factor to numeric for plotting
hourly_mean$HourNum <- as.numeric(as.character(hourly_mean$Hour))

# 
plot(hourly_mean$HourNum, hourly_mean$mean_zone1, type = "l", col = "blue",
     xlab = "Hour of Day", ylab = "Mean Power Consumption (kWh)",
     ylim = range(hourly_mean[, c("mean_zone1","mean_zone2","mean_zone3")]),
     main = "Average Daily Power Consumption Pattern")
lines(hourly_mean$HourNum, hourly_mean$mean_zone2, col = "green")
lines(hourly_mean$HourNum, hourly_mean$mean_zone3, col = "salmon")
legend("topright", legend = c("Zone1","Zone2","Zone3"), col = c("blue","green","salmon"), lty = 1)



```

```{r}
#| echo: false
#| fig.show: hold
#| fig.cap: "aggregate temperature by month for each zone"

####### aggregate temperature for plotting, monthly

monthly.agg <- tetouan %>%
  group_by(Month) %>%
  summarise(avg.temp = mean(Temperature),
            avg.zone1 = mean(Zone1),
            avg.zone2 = mean(Zone2),
            avg.zone3 = mean(Zone3))

plot(monthly.agg$avg.temp, monthly.agg$avg.zone1, type = "b", pch = 19,
     xlab = "Avg Monthly Temperature (°C)", ylab = "Average Monthly Power",
     main = "Monthly Average Power vs Temperature (Zone 1)",  col = "blue")
text(monthly.agg$avg.temp, monthly.agg$avg.zone1, labels = monthly.agg$Month, pos = 3, cex = 0.7)

plot(monthly.agg$avg.temp, monthly.agg$avg.zone2, type = "b", pch = 19,
     xlab = "Avg Monthly Temperature (°C)", ylab = "Average Monthly Power",
     main = "Monthly Average Power vs Temperature (Zone 2)", col = "green")
text(monthly.agg$avg.temp, monthly.agg$avg.zone2, labels = monthly.agg$Month, pos = 3, cex = 0.7)

plot(monthly.agg$avg.temp, monthly.agg$avg.zone3, type = "b", pch = 19,
     xlab = "Avg Monthly Temperature (°C)", ylab = "Average Monthly Power",
     main = "Monthly Average Power vs Temperature (Zone 3)", col = "salmon")
text(monthly.agg$avg.temp, monthly.agg$avg.zone3, labels = monthly.agg$Month, pos = 3, cex = 0.7)


```

## Multiple Linear Regression

```{r}
#| echo: false
#| results: hide


################################################################################
##### mlr for zone 1

# model 
zone1lm <- lm(Zone1 ~ Temperature + Humidity + WindSpeed +  
            GeneralDiffuse + DiffuseFlows + Hour + Weekday + Month, data = train)

summary(zone1lm)
pred <- predict(zone1lm, test)

lm.mse1 <- mean((pred - test$Zone1)^2)
print(lm.mse1)
# 6860050
lm.rmse1 <- sqrt(mean((pred - test$Zone1)^2))
print(lm.rmse1)
# 2619.17
lm.adjr1 <- summary(zone1lm)$adj.r.squared
print(lm.adjr1)
#  0.8626 



################################################################################
##### mlr for zone 2

# model 
zone2lm <- lm(Zone2 ~ Temperature + Humidity + WindSpeed +  
                GeneralDiffuse + DiffuseFlows + Hour +Weekday + Month, data = train)

summary(zone2lm)
  
pred <- predict(zone2lm, test)
lm.mse2 <- mean((pred - test$Zone2)^2)
print(lm.mse2)
# 4452233
lm.rmse2 <- sqrt(lm.mse2)
print(lm.rmse2)
# 2110.031
lm.adjr2 <- summary(zone2lm)$adj.r.squared
print(lm.adjr2)
# 0.834 



################################################################################
##### mlr for zone 3

# model 
zone3lm <- lm(Zone3 ~ Temperature + Humidity + WindSpeed +  
                GeneralDiffuse + DiffuseFlows + Hour + Weekday + Month, data = train)

summary(zone3lm)

pred <- predict(zone3lm, test)
lm.mse3 <- mean((pred - test$Zone3)^2)
print(lm.mse3)
# 5225318
lm.rmse3 <- sqrt(lm.mse3)
print(lm.rmse3)
# 2285.895
lm.adjr3 <- summary(zone3lm)$adj.r.squared
print(lm.adjr3)
# 0.8783 


```

```{r}
#| echo: false
#| results: hold

# combined results
lm.results <- data.frame( Zone = c("Zone 1", "Zone 2", "Zone 3"), MSE = c(lm.mse1, lm.mse2, lm.mse3), RMSE = c(lm.rmse1, lm.rmse2, lm.rmse3), R2 = c(lm.adjr1, lm.adjr2, lm.adjr3))

print(lm.results)


```



## Lasso Regression

```{r}
#| echo: false
#| results: hide

xmat <- model.matrix(~ Temperature + Humidity + WindSpeed + GeneralDiffuse + DiffuseFlows + Hour + Weekday + Month, data = tetouan)[, -1]


################################################################################
##### lasso for zone 1

y1 <- tetouan$Zone1

cvfit.lasso1 <- cv.glmnet(xmat, y1, subset = trainind, alpha = 1)
plot(cvfit.lasso1)
lambest.lasso1 <- cvfit.lasso1$lambda.min
lasso.zone1 <- glmnet(xmat, y1, subset = trainind, alpha = 1, lambda = lambest.lasso1)

predlasso1 <- predict(lasso.zone1, s = lambest.lasso1, newx = xmat[testind, ])

lasso.mse1 <- mean((predlasso1 - y1[testind])^2)
lasso.rmse1 <- sqrt(lasso.mse1)
lasso.adjr1 <- 1 - sum((y1[testind] - predlasso1)^2) / sum((y1[testind] - mean(y1[testind]))^2)


print(lasso.mse1)
# 6851830
print(lasso.rmse1)
# 2617.6
print(lasso.adjr1)
# 0.8652688


################################################################################
##### lasso for zone 2

y2 <- tetouan$Zone2

cvfit.lasso2 <- cv.glmnet(xmat, y2, subset = trainind, alpha = 1)
plot(cvfit.lasso2)

lambest.lasso2 <- cvfit.lasso2$lambda.min
lasso.zone2 <- glmnet(xmat, y2, subset = trainind, alpha = 1, lambda = lambest.lasso2)

predlasso2 <- predict(lasso.zone2, s = lambest.lasso2, newx = xmat[testind, ])

lasso.mse2 <- mean((predlasso2 - y2[testind])^2)
lasso.rmse2 <- sqrt(lasso.mse2)
lasso.adjr2 <- 1 - sum((y2[testind] - predlasso2)^2) /sum((y2[testind] - mean(y2[testind]))^2)

print(lasso.mse2)
# 4447962
print(lasso.rmse2)
# 2109.019
print(lasso.adjr2)
# 0.837011



################################################################################
##### lasso for zone 3

y3 <- tetouan$Zone3


cvfit.lasso3 <- cv.glmnet(xmat, y3, subset = trainind, alpha = 1)
plot(cvfit.lasso3)

lambest.lasso3 <- cvfit.lasso3$lambda.min
lasso.zone3 <- glmnet(xmat, y3, subset = trainind, alpha = 1, lambda = lambest.lasso3)

predlasso3 <- predict(lasso.zone3, s = lambest.lasso3, newx = xmat[testind, ])

lasso.mse3 <- mean((predlasso3 - y3[testind])^2)
lasso.rmse3 <- sqrt(lasso.mse3)
lasso.adjr3 <- 1 - sum((y3[testind] - predlasso3)^2) / sum((y3[testind] - mean(y3[testind]))^2)

print(lasso.mse3)
# 5216769
print(lasso.rmse3)
# 2284.025
print(lasso.adjr3)
# 0.879915




```
```{r}
#| echo: false
#| results: hold
#| cache: true

# combined results
lasso.results <- data.frame( Zone = c("Zone 1", "Zone 2", "Zone 3"), MSE = c(lasso.mse1, lasso.mse2, lasso.mse3), RMSE = c(lasso.rmse1, lasso.rmse2, lasso.rmse3),
                             R2 = c(lasso.adjr1, lasso.adjr2, lasso.adjr3))

print(lasso.results)


```

## Ridge Regression

```{r}
#| echo: false
#| results: hide
#| cache: true

xmat <- model.matrix( ~ Temperature + Humidity + WindSpeed + GeneralDiffuse + DiffuseFlows + Hour + Weekday + Month, data = tetouan)[, -1]


################################################################################
##### ridge for zone 1


y1 <- tetouan$Zone1


cvfit.ridge1 <- cv.glmnet(xmat, y1, subset = trainind, alpha = 0)
plot(cvfit.ridge1)

lambest.ridge1 <- cvfit.ridge1$lambda.min
ridge.zone1 <- glmnet(xmat, y1, subset = trainind, alpha = 0, lambda = lambest.ridge1)

predridge1 <- predict(ridge.zone1, s = lambest.ridge1, newx = xmat[testind, ])

ridge.mse1 <- mean((predridge1 - y1[testind])^2)
ridge.rmse1 <- sqrt(ridge.mse1)
ridge.adjr1 <- 1 - sum((y1[testind] - predridge1)^2) /sum((y1[testind] - mean(y1[testind]))^2)


ridge.mse1
# 7108167
ridge.rmse1
# 2666.115
ridge.adjr1
# 0.8602283


################################################################################
##### ridge for zone 2

y2 <- tetouan$Zone2

cvfit.ridge2 <- cv.glmnet(xmat, y2, subset = trainind, alpha = 0)
plot(cvfit.ridge2)

lambest.ridge2 <- cvfit.ridge2$lambda.min
ridge.zone2 <- glmnet(xmat, y2, subset = trainind, alpha = 0, lambda = lambest.ridge2)

predridge2 <- predict(ridge.zone2, s = lambest.ridge2, newx = xmat[testind, ])

ridge.mse2 <- mean((predridge2 - y2[testind])^2)
ridge.rmse2 <- sqrt(ridge.mse2)
ridge.adjr2 <- 1 - sum((y2[testind] - predridge2)^2) /sum((y2[testind] - mean(y2[testind]))^2)


ridge.mse2
# 4589123
ridge.rmse2
#  2142.224
ridge.adjr2
# 0.8318384



################################################################################
##### ridge for zone 3

y3 <- tetouan$Zone3

cvfit.ridge3 <- cv.glmnet(xmat, y3, subset = trainind, alpha = 0)
plot(cvfit.ridge3)

lambest.ridge3 <- cvfit.ridge3$lambda.min
ridge.zone3 <- glmnet(xmat, y3, subset = trainind, alpha = 0, lambda = lambest.ridge3)

predridge3 <- predict(ridge.zone3, s = lambest.ridge3, newx = xmat[testind, ])

ridge.mse3 <- mean((predridge3 - y3[testind])^2)
ridge.rmse3 <- sqrt(ridge.mse3)
ridge.adjr3 <- 1 - sum((y3[testind] - predridge3)^2) / sum((y3[testind] - mean(y3[testind]))^2)


ridge.mse3
# 5305718
ridge.rmse3
# 2303.415
ridge.adjr3
# 0.8778675


```

```{r}
#| echo: false
#| results: hold


# results comparison
ridge.results <- data.frame(Zone = c("Zone 1", "Zone 2", "Zone 3"),MSE = c(ridge.mse1, ridge.mse2, ridge.mse3), RMSE = c(ridge.rmse1, ridge.rmse2, ridge.rmse3), R2 = c(ridge.adjr1, ridge.adjr2, ridge.adjr3))

print(ridge.results)


```


## Random Forest

```{r}
#| echo: false
#| results: hide
#| cache: true

ctrl <- trainControl(method = "cv", number = 3)  

################################################################################
##### random forest for zone 1

rfZone1 <- randomForest(Zone1 ~ Month + Hour + Weekday + Temperature + Humidity + WindSpeed + GeneralDiffuse + DiffuseFlows, data = train, ntree = 100, importance = TRUE)

# prediction
predrf1 <- predict(rfZone1, test)


rf.mse1 <- mean((predrf1 - test$Zone1)^2)
rf.mse1
# 1367986

rf.rmse1 <- sqrt(rf.mse1)
rf.rmse1
# 1169.609

y1 <- test$Zone1
yhat1 <- predrf1
rf.adjr1 <- 1 - sum((y1 - yhat1)^2) / sum((y1 - mean(y1))^2)
rf.adjr1
# 0.9731005



################################################################################
##### random forest for zone 2

rfZone2 <- randomForest(Zone2 ~ Month + Hour + Weekday + Temperature + Humidity + WindSpeed + GeneralDiffuse + DiffuseFlows, data = train, ntree = 100, importance = TRUE)

# prediction
predrf2 <- predict(rfZone2, test)

rf.mse2 <- mean((predrf2 - test$Zone2)^2)
rf.mse2
# 753378.6

rf.rmse2 <- sqrt(rf.mse2)
rf.rmse2
# 867.9738

y2 <- test$Zone2
yhat2 <- predrf2
rf.adjr2 <- 1 - sum((y2 - yhat2)^2) / sum((y2 - mean(y2))^2)
rf.adjr2
# 0.9723936



################################################################################
##### random forest zone 3

rfZone3 <- randomForest(Zone3 ~ Month + Hour + Weekday + Temperature + Humidity + WindSpeed + GeneralDiffuse + DiffuseFlows, data = train, ntree = 100, importance = TRUE)

# prediction
predrf3 <- predict(rfZone3, test)

rf.mse3 <- mean((predrf3 - test$Zone3)^2)
rf.mse3
# 748996.2

rf.rmse3 <- sqrt(rf.mse3)
rf.rmse3
# 865.4457

y3 <- test$Zone3
yhat3 <- predrf3
rf.adjr3 <- 1 - sum((y3 - yhat3)^2) / sum((y3 - mean(y3))^2)
rf.adjr3
# 0.9827588




# importance plots
varImpPlot(rfZone1, main = "Zone 1")
varImpPlot(rfZone2, main = "Zone 2")
varImpPlot(rfZone3, main = "Zone 3")

plot(rfZone1$mse, type="l", lwd=2,
     xlab="Number of Trees",
     ylab="OOB MSE",
     main="OOB Error - Zone 1")


```


```{r}
#| echo: false
#| results: hold


# results comparison
rf.results <- data.frame(Zone = c("Zone 1", "Zone 2", "Zone 3"), MSE = c(rf.mse1, rf.mse2, rf.mse3), RMSE = c(rf.rmse1, rf.rmse2, rf.rmse3), R2 = c(rf.adjr1, rf.adjr2, rf.adjr3))

print(rf.results)


```



## Gradient Boosting

```{r}
#| echo: false
#| results: hide
#| cache: true
#| message: false
#| warning: false

# revert hour, improves the results of gradient boosting 
train$HourNum <- as.numeric(as.character(train$Hour))  
test$HourNum  <- as.numeric(as.character(test$Hour))

predictors.xgb <- c("Temperature", "Humidity", "WindSpeed", 
                    "GeneralDiffuse", "DiffuseFlows", 
                    "HourNum", "Weekday", "Month")

train.x <- model.matrix(~ . - 1, data = train[, predictors.xgb])
test.x  <- model.matrix(~ . - 1, data = test[, predictors.xgb])

dtrain1 <- xgb.DMatrix(data = train.x, label = train$Zone1)
dtest1  <- xgb.DMatrix(data = test.x, label = test$Zone1)


# zone 1
zone1cv <- xgb.cv(data = dtrain1, nrounds = 1000, nfold = 10,  max_depth = 5, 
                  verbose = 0, eta = 0.6, early_stopping_rounds = 20)


cvlog <- zone1cv$evaluation_log
best.nrounds1 <- which.min(cvlog$test_rmse_mean)

zone1xgmodel <- xgboost(data = train.x, label = train$Zone1, nrounds = best.nrounds1,
                  verbose = 0, params = list(max_depth = 5, eta = 0.6))

predxg1 <- predict(zone1xgmodel, test.x)
xg.mse1 <- mean((predxg1 - test$Zone1)^2)
xg.mse1
# 1179826
xg.rmse1 <- sqrt(xg.mse1)
xg.rmse1
# 1086.198




# zone 2

dtrain2 <- xgb.DMatrix(data = train.x, label = train$Zone2)
dtest2 <- xgb.DMatrix(data = test.x, label = test$Zone2)

zone2cv <- xgb.cv(data = dtrain2, nrounds = 1000, nfold = 10, verbose = 0,
                  max_depth = 5, eta = 0.5, early_stopping_rounds = 20)


cvlog2 <- zone2cv$evaluation_log
best.nrounds2 <- which.min(cvlog2$test_rmse_mean)

zone2xgmodel <- xgboost(data = train.x, label = train$Zone2, nrounds = best.nrounds2,
                        verbose = 0, params = list(max_depth = 5, eta = 0.5))

predxg2 <- predict(zone2xgmodel, test.x)
xg.mse2 <- mean((predxg2 - test$Zone2)^2)
xg.mse2
# 605373
xg.rmse2 <- sqrt(xg.mse2)
xg.rmse2
# 778.0572
  



# zone 3

dtrain3 <- xgb.DMatrix(data = train.x, label = train$Zone3)
dtest3 <- xgb.DMatrix(data = test.x, label = test$Zone3)

zone3cv <- xgb.cv(data = dtrain3, nrounds = 1000, nfold = 10, verbose = 0,
                  max_depth = 5, eta = 0.5, early_stopping_rounds = 20)


cvlog3 <- zone3cv$evaluation_log
best.nrounds3 <- which.min(cvlog3$test_rmse_mean)

zone3xgmodel <- xgboost(data = train.x, label = train$Zone3, nrounds = best.nrounds3,
                        verbose = 0, params = list(max_depth = 5, eta = 0.5))

predxg3 <- predict(zone3xgmodel, test.x)
xg.mse3 <- mean((predxg3 - test$Zone3)^2)
xg.mse3
# 606962.7 
xg.rmse3 <- sqrt(xg.mse3)
xg.rmse3
# 779.0781


# feature importance for zone 1
importance1 <- xgb.importance(model = zone1xgmodel, feature_names = colnames(train.x))
xgb.plot.importance(importance1, main = "Zone 1 - Feature Importance")

# feature importance for zone 2
importance2 <- xgb.importance(model = zone2xgmodel, feature_names = colnames(train.x))
xgb.plot.importance(importance2, main = "Zone 2 - Feature Importance")

# feature importance for zone 3
importance3 <- xgb.importance(model = zone3xgmodel, feature_names = colnames(train.x))
xgb.plot.importance(importance3, main = "Zone 3 - Feature Importance")



```




```{r}
#| echo: false
#| results: hold


# results comparison
xg.results <- data.frame(Zone = c("Zone 1", "Zone 2", "Zone 3"), MSE = c(xg.mse1, xg.mse2, xg.mse3), RMSE = c(xg.rmse1, xg.rmse2, xg.rmse3))

print(xg.results)


```


